---
title: "IPI Student assistant tasks"
format:
  html:
    df_print: paged
    number-sections: true
    embed-resources: true
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DeclareDesign)
library(knitr)
```

**Instructions**

- Please complete these tasks within _one hour_ of receipt of our email. 

- We strongly encourage you to use `.Rmd` or `.qmd` and to share back both the code and the compiled file as html or pdf. If using other methods please return both code and output.  



# Interpretation (10 mins)

```{r, echo = FALSE}
library(DeclareDesign)

data <- 
  fabricate(N = 100, 
            X1 = simple_ra(N), 
            X2 = block_ra(blocks = X1),
            Y = runif(N) < .25*X1 + .5*X1 * X2)

lm_robust(Y~X1*X2, data = data) |> tidy() |>
  kable(digits = 3)
```

Half the units are assigned to X1 = 0 and half to X1 = 1; half o each of these are assigned to X2=0 and half to X2 = 1.

Using the the coefficients from this regression please indicate:

1. The estimated effect of X1 on Y when X2 = 0
2. The estimated effect of X1 on Y when X2 = 1
3. The average effect of X1 on Y


# Editing (10 mins)

Please read this text and suggest improvements.

<blockquote>
  "In the last twelve month's (See Masters 2022) there have been  <br> less protests about the conflict then any time before <span style="color: white;">1968</span> <br>
  this has lead some to wonder what affect that <br> might have on future stability."
</blockquote>


#  Data management  (40 mins)

You have been given 2 datasets.
 
*  [Dataset *A*](A_CSES.csv) ("A_CSES.csv") is a short version of 
 the Comparative Study of Electoral Systems (CSES) survey wave 2. 
 Each observation in this dataset refers to an individual in a
 certain year in a specific country. 

* [Dataset *B*](B_QoG.xlsx) ("B_QoG.xlsx") is a short version of 
 the Quality of Government (QoG) data. 
 Each observation in this dataset refers to a certain year in 
 a specific country. 

Dataset *A* includes the following variables:

| Variable name | Variable label          | 
|--------|-------------------------|
| B1004         |Country-year code |
| B1006_NAM     | Country name|
| B1008         | Year|
| B2001         | Age (number of years) |
| B2002         |Gender (1=male; 2=female) |
| B2005         | Union membership (1=is a member; 2=is not a member)
| B2020         | Income in household (1=low, 5=high)|
| B3014         | It matters who people vote for (1 = it doesn't matter; 5 = it matters a lot)|
 
 
Dataset *B* includes the following variables:
 
| Variable name | Variable label        | 
|---------------|-----------------------|
| cname          | Country name | 
| year           | Year of measurement| 
| gle_cgdpc      | GDP per capita (in current prices)| 
| p_polity2      | Polity score (measure of democracy) | 
| undp_hdi       | UNDP's Human Development Index | 



## Read data

Read both datasets using the needed functions for
their format. 
     Note that dataset *A* is a .csv (Comma-Separated Value) file, 
     while dataset *B* is an .xlsx (Excel) file.


## Clean data

 Compare the list of unique countries that are present in
     datasets *A* and *B*. 
     There are 5 countries that are named differently between these 2
     datasets. Find these countries and then rename those in dataset
     *B* so that they match the names used in dataset *A*.

## Merge data

Merge dataset *B* into dataset *A* using country and year to match
     observations. The dataset resulting from the merging procedure
     can be called `merged_df`. 
     Please perform the merging so that the resulting dataset, has the same number of rows as dataset *A*. 

#  Descriptive statistics 

 From this point onward, please continue your work only with the merged dataset 
  `merged_df`. (We also provide this dataset [here](merged_df.csv) in case you have not merged properly)

## Summary statistics

For each country-year pair in the merged dataset, please 
     compute the percentage of respondents who report being 
     members of a union, as well as the GDP per capita recorded
     in that country-year. 
     
Store this resulting country-year data as a data frame in   a new object called `summary_df`. 

## Display table

Display the first rows of `summary_df`  as a table.


#  Analysis 

The next questions use `summary_df`. If you did not create `summary_df` successfully you can access it [here](summary_df.csv). 

<!-- ## Compute correlations -->

<!-- Compute the correlation across country-years between the percentage of  -->
<!--      respondents who are members of a labor union and the GDP per -->
<!--      capita.  -->


## Scatterplot

Please produce a scatterplot of the relationship between union membership and GDP per capita.
Plot union membership on the X-axis, and GDP per capita on the Y-axis.


<!-- ## Scatterplot 2 -->

<!-- Please create a new version of the same scatterplot as above, -->
<!--      but here give a different color to the points if the year of  -->
<!--      the observation is 2004. All other years will have the same  -->
<!--      color on the plot, except the countries that were measured in -->
<!--      2004. -->



<!-- ## Regression -->

<!-- Run an OLS regression of union membership on year. Display your output. -->



Note: you may not be able to finish everything; that's OK. If stuck for time indicate briefly the steps you would take.